---
title: 'Reproducible Research: Peer Assessment 1'
output:
  pdf_document: default
  html_document:
    keep_md: yes
---


## Loading and preprocessing the data
```{r}
require(dplyr)
str(df.raw <- unz(
  'activity.zip'
  ,'activity.csv'
  ) %>%
  read.csv(  
    colClasses=c(
      'integer'
      ,'Date'
      ,'integer'
      )
    ) %>%
  mutate(
    weekday = factor(
      weekdays(date, abbreviate=TRUE)
      ,levels=c('Sun','Mon','Tue','Wed','Thu','Fri','Sat')
      )
    ,workweek = factor(ifelse(
      weekday %in% c('Sat', 'Sun')
      ,'weekend'
      ,'weekday'
      ))
    ,date.int = as.integer(date)
    ))
summary(df.raw)
hist(log1p(df.raw$steps), breaks='FD')
hist(df.raw$date, breaks='day')
hist(df.raw$interval, breaks='FD')
df.raw$weekday %>%
  table %>%
  rev %>%
  barplot(
    horiz=TRUE
    ,main='Day of Week Frequency'
    )
df.raw$workweek %>%
  table %>%
  sort %>%
  barplot(
    horiz=TRUE
    ,main='Weekend Frequency'
    )
```


## What is mean total number of steps taken per day?
```{r}
str(df.day <- df.raw %>%
  group_by(date) %>%
  summarize(
    steps_mean = mean(steps, na.rm=TRUE)
    ,steps_median = median(steps, na.rm=TRUE)
    ,steps_total = sum(steps, na.rm=TRUE)
    ) %>%
  arrange(date))
summary(df.day)
hist(
  df.day$steps_total
  ,breaks=30
  ,main='Histogram of Total Steps per Day\nNo Imputation'
  )
print(steps.per.day.mean <- mean(df.day$steps_total, na.rm=TRUE))
print(steps.per.day.median <- median(df.day$steps_total, na.rm=TRUE))
```
  * Mean number of steps per day prior to imputation was `r steps.per.day.mean`
  * Median number of steps per day prior to imputation was `r steps.per.day.median`


## What is the average daily activity pattern?
```{r}
str(df.interval <- df.raw %>%
  group_by(interval) %>%
  summarize(
    steps_mean = mean(steps, na.rm=TRUE)
    ,steps_median = median(steps, na.rm=TRUE)
    ,steps_total = sum(steps, na.rm=TRUE)
    ) %>%
  arrange(interval))
with(df.interval,plot(
  x=interval
  ,y=steps_mean
  ,type='l'
  ,xlab='5-minute interval'
  ,ylab='Mean Steps Taken'
  ,main='Pattern of steps by time interval'
  ))
print(interval.max.steps <- with(df.interval, interval[which.max(steps_mean)]))
abline(v=interval.max.steps,lty=3,col='red')
```
  * The 5-minute interval with the highest average steps taken was interval `r interval.max.steps`


## Imputing missing values
  * There are `r sum(is.na(df.raw$steps))` rows with missing values.
  * Since only one variable has missing values, imputation was done as a classic supervised learning problem.
  * A Random Forest algorithm was used to learn from the observed step counts and predict the missing ones.
```{r}
require(randomForestSRC)

rf.fit <- rfsrc(
  steps ~ .
  ,data=df.raw %>% filter(!is.na(steps)) %>% select(-date)
  ,ntree=1000
  ,mtry=2
  )
print(rf.fit)
plot(rf.fit)

print(rf.predict <- rf.fit %>%
  predict(
    newdata=df.raw %>% filter(is.na(steps)) %>% select(-date, -steps)
    ))
str(df.impute <- df.raw %>%
  mutate(
    steps.filled = steps
    ,imputed = is.na(steps)
    ))
mean(df.impute$steps.filled[df.impute$imputed] <- rf.predict$predicted)
summary(df.impute)
```


### Now repeat the earlier day analysis of steps per day
```{r}
str(df.day.impute <- df.impute %>%
  group_by(date) %>%
  summarize(
    steps_mean = mean(steps.filled)
    ,steps_median = median(steps.filled)
    ,steps_total = sum(steps.filled)
    ) %>%
  arrange(date))
summary(df.day.impute)
hist(
  df.day.impute$steps_total
  ,breaks=30
  ,main='Histogram of Total Steps per Day\nWith Imputation'
  )
print(steps.per.day.impute.mean <- mean(df.day.impute$steps_total, na.rm=TRUE))
print(steps.per.day.impute.median <- median(df.day.impute$steps_total, na.rm=TRUE))
```
  * Mean number of steps per day after imputation was `r steps.per.day.impute.mean`
  * Median number of steps per day after imputation was `r steps.per.day.impute.median`
  * Imputation increased the mean by `r steps.per.day.impute.mean - steps.per.day.mean` steps
  * Imputation increased the median by `r steps.per.day.impute.median - steps.per.day.median` steps
  * The summary statistics after imputation are noticeably higher than those prior to imputation.


## Are there differences in activity patterns between weekdays and weekends?
```{r}
require(ggplot2)
require(scales)
str(df.weekday <- df.impute %>%
  group_by(workweek, interval) %>%
  summarize(
    steps_mean = mean(steps.filled)
    ))
plt.weekday <- df.weekday %>%
  ggplot(aes(
    x=interval
    ,y=steps_mean
    ,color=workweek
    )) +
  facet_wrap(~workweek, ncol=1) +
  geom_line(size=1.5) +
  scale_x_continuous('Interval') +
  scale_y_continuous('Average Number of Steps') +
  scale_color_discrete(guide='none') +
  theme_light()
plot(plt.weekday)
```
  * Weekend step patterns do not have as large of spike around interval `r interval.max.steps`
  * Weekend step patterns rise above 100 steps per interval in more parts of the day.
